---
title: "One decision, many ontologies: a carriage-road path from Camden to the Snow Bowl"
author: "Cory Whitney"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, fig.width = 9, fig.height = 6, dpi = 120)
knitr::opts_knit$set(root.dir = normalizePath(".."))   # run from the project root
```

```{r setup2, include=FALSE}
library(ggplot2)
for (f in list.files("R", full.names = TRUE)) source(f)
stakeholders <- load_stakeholders("stakeholders")
rd <- function(f) read.csv(f, stringsAsFactors = FALSE)
img <- function(...) knitr::include_graphics(normalizePath(c(...)), rel_path = FALSE)
```

> **Status: a scaffold with placeholder inputs.** Every input range in the stakeholder tables is a
> placeholder invented so the pipeline runs, except where a table row says it is cited. The route
> geometry (length, parcels, climb) is measured from public data. Read the results as the *shape* of
> the problem, not as evidence.

## 1. The decision

Should a separated carriage-road path be built between downtown Camden and the Camden Snow Bowl,
instead of relying on Hosmer Pond Road? Each stakeholder group sees the same physical change (a few
hectares of corridor, more trips, some trees) through its own **ontology**: what exists, what causes
what, and what counts as good or bad. Each ontology is a causal diagram (DAG), an input table and an
outcome function; they are combined in one `decisionSupport` Monte Carlo model.

The Snow Bowl is **not** a separate party. It is a Town-owned special revenue fund run by Parks and
Recreation, kept apart from the General Fund (Camden FY2026 municipal budget message). So the Town and
the Snow Bowl form one ontology and one player, with two budget pockets.

```{r ontologies}
data.frame(Group = vapply(stakeholders, `[[`, "", "label"),
           `Sees the road as` = vapply(stakeholders, `[[`, "", "ontology"),
           Values = vapply(stakeholders, function(s) nrow(s$values), 1L), check.names = FALSE, row.names = NULL)
```

## 2. The geography (measured)

Routes are cheapest paths on OpenStreetMap roads and trails (`gis/02_routes.R`); parcels are from the
Maine GeoLibrary and elevation from USGS 3DEP (`gis/03_parcels_elevation.R`).

```{r map, fig.height=6.5}
plot_routes_map()
```

```{r gis-tables}
rs <- rd("gis/outputs/route_summary.csv"); ps <- rd("gis/outputs/parcel_summary.csv")
merge(rs[, c("route", "length_km", "pct_hosmer_pond_road", "straight_line_km")],
      ps[, c("route", "parcels_touched", "private_parcels_touched", "public_parcels_touched")], by = "route")
```

"Parcels touched" are tax parcels within 15 m of the route line. Public means the owner name contains
Town of, State of, Land Trust or Conservation; private owner names were never downloaded. Parcel data
are submitted by towns on an unscheduled basis and may be old.

```{r elevation, fig.height=4}
plot_elevation()
rd("gis/outputs/grade_summary.csv")
```

## 3. Stage 1: what each group gains or loses

The bubble matrix shows, for each group's own values, the median change against the status quo (blue
gain, red loss), sized by magnitude on that value's own scale and faded by uncertainty about the sign.

```{r bubble, fig.width=10, fig.height=13}
summ <- rd("outputs/liability_protected/value_summary.csv")
bubble_matrix(summ, stakeholders,
              options = c("shoulders", "carriage_path", "path_donated", "path_cost_sharing", "partial_path"),
              title = "Who gains and who loses, in each group's own terms")
```

```{r heatmap, echo=FALSE, out.width="100%"}
img("outputs/liability_protected/welfare_heatmap.png")
```

With the placeholders every option makes some group worse off, so there is no neutral choice: a
recommendation is a distributional choice. The all-groups Pareto views are in `outputs/`.

## 4. The Town's two pockets

```{r pockets, fig.height=5}
plot_budget_pockets()
```

Because the Snow Bowl fund belongs to the Town, the question is not "who among the Town and the Snow
Bowl pays" but whether the margin the path earns in the Snow Bowl fund can be **earmarked** to the path's
cost, and what the General Fund must cover. The Snow Bowl fund itself runs close to break even
(FY2026 budget message), so the margin is a new inflow, not a reserve.

## 5. Stage 2: a game of three players

Players: the Town (including the Snow Bowl), landowners (as one collective player) and local business.

```{r game-tables}
pick <- function(f) { o <- rd(f); o$nash_prob <- round(o$nash_prob, 2); o[1:6, c("label", "nash_prob", "nash_at_mean")] }
cat("Voluntary contributions (protected liability scenario)")
pick("outputs/liability_protected/game_voluntary_outcomes.csv")
cat("Binding cost-sharing agreement")
pick("outputs/liability_protected/game_binding_outcomes.csv")
```

```{r game-plots, echo=FALSE, out.width="100%"}
img(c("outputs/liability_protected/game_voluntary_nash_probability.png",
                          "outputs/mechanism_comparison.png", "outputs/liability_protected/mechanism_design_grid.png"))
```

### What changed when the Snow Bowl joined the Town

An earlier version treated the Snow Bowl as a separate player. With the Snow Bowl's gains counted in the
Town's own payoff, the Town no longer loses from the path at the median, and a path becomes a stable
outcome in most simulated worlds. The earlier finding that "the Town is the bottleneck" came from leaving
those gains out of its payoff. This is a modelling lesson, not yet a result: the Snow Bowl inputs are
placeholders.

## 6. Liability scenarios

Whether paying hosts for an easement counts as "consideration" under Maine's recreational use
statute ([14 MRSA 159-A](https://legislature.maine.gov/statutes/14/title14sec159-A.html)) is a legal
question that needs a Maine attorney. Both cases run as scenarios.

```{r liability, echo=FALSE, out.width="90%"}
img("outputs/liability_comparison.png")
```

## 7. Where the numbers come from

* **Placeholders:** all ranges unless marked. See `inputs/all_inputs.csv`.
* **Cited:** carbon price range in the NPV Pareto view (Rennert et al. 2022). Methods behind the other
  exchange rates are cited in `R/exchange_rates.R`; their per-unit numbers remain placeholders.
* **Measured:** route lengths, parcel counts, climb (GIS).
* **Verified fact:** the Snow Bowl is a Town special revenue fund outside the General Fund.

```{r refs, echo=FALSE, results="asis"}
cat("**References** (checked against Crossref, 2026-10-05)\n\n")
cat("* Rennert, K., Errickson, F., Prest, B. C., Rennels, L., Newell, R. G., Pizer, W. (2022). Comprehensive evidence implies a higher social cost of CO2. *Nature* 610, 687-692. https://doi.org/10.1038/s41586-022-05224-9\n")
cat("* De Neve, J.-E., Clark, A. E., Krekel, C., Layard, R., O'Donnell, G. (2020). Taking a wellbeing years approach to policy choice. *BMJ*, m3853. https://doi.org/10.1136/bmj.m3853\n")
cat("* Costanza, R., de Groot, R., Sutton, P., van der Ploeg, S., Anderson, S. J., Kubiszewski, I. (2014). Changes in the global value of ecosystem services. *Global Environmental Change* 26, 152-158. https://doi.org/10.1016/j.gloenvcha.2014.04.002\n")
cat("* Arrow, K., Cropper, M., Gollier, C., Groom, B., Heal, G., Newell, R. (2013). Determining benefits and costs for future generations. *Science* 341, 349-350. https://doi.org/10.1126/science.1235665\n")
cat("* Drupp, M. A., Freeman, M. C., Groom, B., Nesje, F. (2018). Discounting disentangled. *American Economic Journal: Economic Policy* 10, 109-134. https://doi.org/10.1257/pol.20160240\n")
```

Map data: (c) OpenStreetMap contributors (ODbL). Parcels: Maine GeoLibrary. Elevation: USGS 3DEP.
